Supplementary methods

### 2020-08-28 text/supplemental_methods.Rmd

@@

In order to understand how the diagnosis of cancer changes the expectation for
longevity of patients, we first modeled the longevity of the overall population
of the U.S. based on CDC life tables. Most TCGA cases are diagnosed late in
life, so we fit a model that emphasizes accuracy in the right tail of the
distribution. We modeled baseline longevity using a risk rate function:

>>>>

### Selection of pathways

To understand the molecular mechanism of cancer at a pathway level, we used
Reactome (https://reactome.org/), a knowledge-based pathway database. The
mapping files of ENSEMBL genes to pathways, pathway hierarchy relationships, and
complexes to top pathways were downloaded from
https://reactome.org/download-data. Using these data, we imposed pathway
criteria to define basic cellular processes and biochemical pathways: (1)
human-derived pathways (\it{“HSA”}) (2) limited to grandchild node for each
parent pathway (_e.g._ ‘Beta-catenin independent WNT signaling’ in ‘Signal
Transduction’) (3) exclusion of pathways in the parent pathway: "Disease",
"Muscle contraction", and "Reproduction" or pathway names that include any of
the following keywords: "disease", "defect", "cancer", "oncogenic", "disorder",
"mutant", "loss", "infection", "bacteria", or "listeria". While some of the
excluded pathways have been shown to play an important role in cancer, they are
highly specialized (_e.g._ "PI3K/AKT Signaling in Cancer"). Additionally, for
most of the excluded pathways, a neutral version pathway of the pathway exists
(_e.g._ "PIP3 activates Akt signaling"). Finally, we mapped the 18,577
Ensembl IDs from the TCGA dataset to the highly selected Reactome pathways. This
operation produced a lookup table that consisted of 377 pathways mapped to 8,940
genes.

### Filtering genes

We filtered likely erroneous mutations due to transcription coupled repair. Our
approach was to determine the status of each gene (_i.e._ expressed or not
expressed) in each tissue type in order to exclude gene mutations low expressed
enes. To do this, we obtained the TCGA RNA sequencing data adjusted for batch
effect dataset (https://pancanatlas.xenahubs.net). Using the data, we removed
the genes and tumor samples that were not included in our analysis, grouped the
tumor samples by tissue type and computed the mean expression value for each
gene. A minimum threshold of 10 transcripts per million was set for expressed
genes based on an inflection point observed when plotting the mean expression
values of genes ranked by expression in each tissue type. Genes that did not
meet this threshold were considered not expressed. This operation produced a
lookup table for gene expression status in each tumor sample for 18,127 genes.

### Class and stage-specific Enrichment Calculations

We calculated the enrichment of pathways in one set of tumors as the relative
fraction of tumors (with estimated uncertainty) with a mutated gene in the
pathway to all tumors, inclusive of the category of interest. To do this
operation, we use the beta distribution to permute a posterior distribution on
the fraction of tumors with a pathway mutated for each category based on the
observed set, and compared this to the posterior obtained from the full set of
tumors (all tumors) as the distribution of differences between all permuted
samples. We considered a pathway enriched if the 95% range of credible
differences thus obtained excludes 0 and the mean credible difference was
greater than or equal to 30% enrichment, which excludes a large number of small
differences that are not likely to be biologically relevant.

@@

We used Bayesian inference to explore the relative impact of cancer diagnosis on
survival at the time of diagnosis. Our model assumes that at birth each person
has a small initial probability per unit time to die, and that this probability
grows exponentially in time at some baseline rate that matches the observed CDC
data. The rate of growth in risk is assumed to jump to a higher level at
diagnosis, which we take as a proxy for onset. This represents an accelerated
aging type model. To fit the model, we constrain the baseline risks using the
available CDC life tables (ftp://ftp.cdc.gov/pub/Health_Statistics/NCHS/Publications/NVSR/68_07/). We also constrain the cancer specific multiplier
using the longevity data within the cancer tumor dataset. Our model estimates
the effect of cancer type and class-specific cancer effects independently,
resulting in a cancer and class specific estimate of the aging rate multiplier.

Most TCGA cases are diagnosed late in life, so we fit a model that emphasizes
accuracy in the right tail of the distribution. We modeled baseline longevity
using a risk rate function:
